Exploring the Role of Feedback in Machine Learning Jupyter Notebooks

### 2024-06-28 report/emse24.tex
feat(report): emse24 updates from Luis & Diomidis

The MLDL is fundamentally iterative, closely resembling Agile software engineering principles which emphasize incorporating early feedback and working in small, manageable iterations. Similarly, CRISP-DM---the dominant process model in data mining---outlines a cyclical process that evolves from understanding the business objectives and data, through data preparation, modeling, and evaluation, to deployment and monitoring. Both systems advocate for a dynamic, iterative approach that enhances adaptability and effectiveness in dealing with complex, changing systems like those encountered in ML projects~\cite{betz2018managing,martinez-plumed2021crisp-dm}.

Computational notebooks---specifically Jupyter notebooks~\footnote{https://jupyter.org/}---have been ubiquitously adopted by the ML community for developing ML enabled systems~\cite{pimentel2019large-scale,quaranta2021kgtorrent,psallidas2019data,perkel2018why}. Jupyter notebooks provide an interactive computing environment that allow users to break complex code into smaller, more manageable chunks. Each notebook is composed of a series of code cells, which can be executed independently in a Read-Eval-Print Loop (REPL) style of development. This REPL approach allows users to write a piece of code, run it, and immediately see the results thus facilitating an incremental and iterative style of software development. This modular structure of notebooks provides users the flexibility to experiment with different approaches and algorithms. Users can test hypotheses, debug issues, and make incremental changes with immediate feedback. The ability to visualize data and results inline and document the process in a narrative form further enhances their utility~\cite{kery2018story,head2019managing,rule2018exploration,chattopadhyay2020whats}.

Feedback is crucial for making important decisions throughout the MLDL. Practitioners rely on the output of cells to determine the appropriate data preprocessing techniques, create new features to boost model performance, and select specific hyperparameters. However, these decisions are often based on the conditions present when the model was initially created. Visual outputs in particular, are open to interpretation~\cite{heer2010tour}. The viewer's background and expertise significantly influence how they understand and draw conclusions from visual data. 
For instance, interpreting a histogram requires knowledge of various distribution types, which someone without statistical training might not possess. Additionally, physical limitations such as color blindness can impede accurate interpretation if visualizations are not designed with accessibility in mind. As data, personnel, and system environments inevitably change, undocumented decisions can lead to accumulation of technical debt. This not only prolongs time required for debugging and troubleshooting, but also incurs additional costs and resource expenditures~\cite{sculley2015hidden,amershi2019software,sambasivan2021everyone}.

@@

To understand the role of feedback when developing ML pipelines in Jupyter notebooks, we conduct an empirical study of the outputs of code cells in Jupyter notebooks. Next, we conduct an empirical study of Python assertions written in Jupyter notebooks to understand how they are used when developing ML pipelines in notebooks. Finally, we investigate the intersection of outputs and assertions by conducting an empirical analysis of assertions that document decisions made from the outputs of code cells. In particular, we focus on assertions derived from visualisations since they pose a higher risk of technical debt.

@@

Initially, Sara loads a dataset and uses basic descriptive statistics to understand its distribution. As she executes a cell that calculates and displays the mean and standard deviation of each feature, she notices unusually high variance in one of the features. This insight, derived directly from the output of the notebook cell, prompts her to apply normalization to ensure better model performance.

@@

Throughout this process, Sara heavily relies on the outputs of various code cells, not just for error correction but also for making strategic decisions about model architecture, such as adding layers or changing activation functions. At one point, after noticing overfitting from a cell outputting validation loss, she decides to incorporate dropout layers, which she can quickly evaluate by rerunning the training process.

@@

Our results from RQ1 reveal that outputs from code cells in Jupyter notebooks are extensively used to make design and implementation decisions during the ML development cycle. Making such decisions however, requires manual validation of the outputs. For instance, we find several instances where practitioners use visualisations to check data distributions and print statements to verify data shapes and types before proceeding with model training. Critical decisions, such as deciding whether to normalize features based on their distributions or adjusting model parameters based on training feedback, rely heavily on these manual checks. Like traditional software systems, an ML pipeline is subject to change due to evolving model requirements, personnel turnover, and changes in the data itself. Such changes can introduce inconsistencies or errors that manual checks might miss, highlighting the need for more automated validation mechanisms.

@@

From the 3M Jupyter notebooks collected in this study, only 30K (9.7\%) were found to utilize assertions. This indicates that while the use of assertions in Jupyter notebooks is still uncommon, it represents an emerging testing practice within the ML community. The relatively low adoption rate suggests that there is significant room for growth and improvement in the integration of assertions into ML development workflows.

@@

Visualizations are an effective tool for interpreting complex ML models and data transformations. However, the manual effort required to validate insights derived from these visualizations can be substantially more compared to text output such as the ones presented in Section~\ref{sec:result-rq2}. We argue that the need for translating visual insights into formal assertions is more pressing. The visual cue and trends observed in visualisations are open to iterpretation, thus the interpretation of the same plot may differ amongst two practitioners. Translating visual insights into assertions ensures taht the knowledge encapsulated in the visualisations along with its interpretation and subsequent decisions made based on this interpretation is preserved and consistently applied throughout the ML lifecycle.

>>>>

The MLDL is fundamentally iterative, closely resembling Agile software engineering principles, which emphasize incorporating early feedback and working in small, manageable iterations~\cite{betz2018managing}. Similarly, CRISP-DM---the dominant process model in data mining---outlines a cyclical process that evolves from understanding the business objectives and data, through data preparation, modeling, and evaluation, to deployment and monitoring~\cite{martinez-plumed2021crisp-dm}. Both systems advocate for a dynamic, iterative approach that enhances adaptability and effectiveness in dealing with complex, changing systems like those encountered in ML projects.

Computational notebooks---specifically Jupyter notebooks\footnote{https://jupyter.org/}---have been ubiquitously adopted by the ML community for developing ML enabled systems~\cite{pimentel2019large-scale,quaranta2021kgtorrent,psallidas2019data,perkel2018why}. Jupyter notebooks provide an interactive computing environment that allow users to break the development of complex workflows into smaller, more manageable code chunks. Each notebook is composed of a series of code cells, which can be executed independently in a read-eval-print loop (REPL) style of interactive software development. This REPL approach allows users to write a piece of code, run it, and immediately see the results thus facilitating an incremental and iterative style of software development. This modular structure of notebooks provides users the flexibility to experiment with different approaches and algorithms. Users can test hypotheses, debug issues, and make incremental changes with immediate feedback. The ability to visualize data and results inline and document the process in a narrative form further enhances their utility~\cite{kery2018story,head2019managing,rule2018exploration,chattopadhyay2020whats}.

Feedback is crucial for making important decisions throughout the MLDL. Practitioners rely on the output of cells to determine the appropriate data preprocessing techniques, detect erroneous assumptions, create new features to boost model performance, and select specific hyperparameters. However, these decisions are often based on the conditions present when the model was initially created. Visual outputs in particular, are open to interpretation~\cite{heer2010tour}. The viewer's background and expertise significantly influence how they understand and draw conclusions from visual data. 
For instance, interpreting a histogram requires knowledge of various distribution types, which someone without statistical training might not possess. Additionally, physical limitations such as color blindness can impede accurate interpretation if visualizations are not designed with accessibility in mind. As data, personnel, and system environments inevitably change, undocumented decisions can lead to accumulation of technical debt. This not only prolongs time required for debugging and troubleshooting, but also incurs additional costs and resource expenditures~\cite{sculley2015hidden,amershi2019software,sambasivan2021everyone}.

@@

To understand the role of feedback when developing ML pipelines in Jupyter notebooks, we conduct a series of  empirical studies. First, we examine the outputs of code cells in Jupyter notebooks. Next, we study Python assertions written in Jupyter notebooks to understand how they are used when developing ML pipelines in notebooks. Finally, we investigate the intersection of outputs and assertions by analysing  assertions that document decisions made from the outputs of code cells. In particular, we focus on assertions derived from visualisations since they pose a higher risk of technical debt.

@@

Initially, Sara loads a dataset and runs basic descriptive statistics to understand its distribution. As she executes a cell that calculates and displays the mean and standard deviation of each feature, she notices unusually high variance in one of the features. This insight, derived directly from the output of the notebook cell, prompts her to apply normalization to ensure better model performance.

@@

Throughout this process, Sara heavily relies on the outputs of various code cells, not just for detecting and fixing erroneous assumptions and code, but also for making strategic decisions about the model's architecture, such as adding layers or changing activation functions. At one point, after noticing overfitting from a cell outputting validation loss, she decides to incorporate dropout layers, which she can quickly evaluate by rerunning the training process.

@@

Our results from RQ1 reveal that outputs from code cells in Jupyter notebooks are extensively used to make design and implementation decisions during the ML development cycle. Making such decisions however, requires manual validation of the outputs. For instance, we find several instances where practitioners use visualisations to check data distributions and print statements to verify data shapes and types before proceeding with model training. Critical decisions, such as deciding whether to normalize features based on their distributions or adjusting model parameters based on training feedback, rely heavily on these manual checks. Like traditional software systems, an ML pipeline is subject to change due to evolving model requirements, personnel turnover, and changes in the data itself. Such changes can introduce inconsistencies or errors that manual checks might miss, highlighting the need for more automated and explicit validation mechanisms.

@@

From the 3 million Jupyter notebooks collected in this study, only 30 thousand (9.7\%) were found to utilize assertions. This indicates that while the use of assertions in Jupyter notebooks is still uncommon, it represents an emerging testing practice within the ML community. The relatively low adoption rate suggests that there is significant room for growth and improvement in the integration of assertions into ML development workflows.

@@

Visualizations are an effective tool for interpreting complex ML models and data transformations. However, the manual effort required to validate insights derived from these visualizations can be substantially more compared to text output such as the ones presented in Section~\ref{sec:result-rq2}. We argue that the need for translating visual insights into formal assertions is more pressing. The visual cue and trends observed in visualisations are open to interpretation, thus the interpretation of the same plot may differ amongst two practitioners. Translating visual insights into assertions ensures that the knowledge encapsulated in the visualisations along with its interpretation and subsequent decisions made based on this interpretation is preserved and consistently applied throughout the ML lifecycle.
